MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators

### 2024-03-22 Article.tex
new format

%\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algpseudocode}

\title{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}
% \shorttitle{MaBoSS for HPC environments}

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1,\authfn{1}]{Adam Šmelko}
\author[2]{Miroslav Kratochvíl}
\author[3,4,5]{Emmanuel Barillot}
% \author[3,4,5]{Laurence Calzone}
\author[3,4,5,\authfn{1}]{Vincent Noël}

\affil[1]{Department of Distributed and Dependable Systems, Charles University, Prague, Czech Republic}
\affil[2]{Luxembourg Centre for Systems Biomedicine, University of Luxembourg, Esch-sur-Alzette, Luxembourg}
\affil[3]{Institut Curie, Université PSL, F-75005, Paris, France}
\affil[4]{INSERM, U900, F-75005, Paris, France}
\affil[5]{Mines ParisTech, Université PSL, F-75005, Paris, France}

\authnote{\authfn{1}smelko@d3s.mff.cuni.cz; vincent.noel@curie.fr}

\papercat{Technical Note}

\runningauthor{Šmelko et al.}

\jvolume{00}
\jnumber{0}
\jyear{2023}

@@

\begin{frontmatter}
\maketitle

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Computational models in systems biology are becoming more important with the advancement of experimental techniques to query the mechanistic details responsible for leading to phenotypes of interest. In particular, Boolean models are well fit to describe the complexity of signaling networks while being simple enough to scale to a very large number of components. With the advance of Boolean model inference techniques, the field is transforming from an artisanal way of building models of moderate size to a more automatized one, leading to very large models. In this context, adapting the simulation software for such increases in complexity is crucial. 
We present two new developments in the continuous time Boolean simulators: MaBoSS.MPI, a parallel implementation of MaBoSS which can exploit the computational power of very large CPU clusters, and MaBoSS.GPU, which can use GPU accelerators to perform these simulations. 
\end {abstract}
%TC:break main
%the command above serves to have a word count for the abstract

\begin{keywords}
Computational Biology | High Performance Computing | Boolean models
\end{keywords}
\end{frontmatter}

@@

The source code of the proposed implementations is publicly available at their respective GitHub repositories\footnote{\url{https://github.com/sysbio-curie/MaBoSS.GPU}} \footnote{\url{https://github.com/sysbio-curie/MaBoSS-env-2.0}}. We also provide the scripts, presented plots, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/sysbio-curie/gigascience24-artifact}}.

@@

\subsubsection{Simulation}

@@

@@

The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer~\footnote{\url{https://www.bsc.es/marenostrum/marenostrum}}.

@@

\section{Availability of source code and requirements}
% flex and bison are linux only :/
\begin{itemize}
	\item Project name: MaBoSS.GPU
	\item Project home page: \url{https://github.com/sysbio-curie/MaBoSS.GPU}
	\item Operating system(s): Platform independent
	\item Programming language: C++, CUDA
	\item Other requirements: Flex, Bison, CMake >= 3.18, Cuda toolkit >= 12.0
	\item License: MIT
\end{itemize}

@@

\section{Data availability}
The scripts, presented plots, data and instructions to reproduce the benchmarks are available on GitHub~\cite{adam_smelko_2024_10841007}.

\subsection{Competing Interests}
The authors declare that they have no competing interests.

\section{Funding}
The research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
% This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773]
The project was partially supported by Charles University, SVV project number~260698.

\section{Acknowledgements}
We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.

% \section*{Bibliography}
\bibliography{bibliography}

%% You can use these special %TC: tags to ignore certain parts of the text.
%TC:ignore
%the command above ignores this section for word count
% \onecolumn
% \newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Supplementary Information %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\captionsetup*{format=largeformat}
%\section{Something about something} \label{note:Note1} 
%Something

%TC:endignore
%the command above ignores this section for word count

>>>>

%\usepackage{blindtext}
\usepackage{algorithm}
\usepackage{algpseudocode}
%%%%

%%%%%=============================================================================%%%%
%%%%  Remarks: This template is provided to aid authors with the preparation
%%%%  of original research articles intended for submission to journals published 
%%%%  by Springer Nature. The guidance has been prepared in partnership with 
%%%%  production teams to conform to Springer Nature technical requirements. 
%%%%  Editorial and presentation requirements differ among journal portfolios and 
%%%%  research disciplines. You may find sections in this template are irrelevant 
%%%%  to your work and are empowered to omit any such section if allowed by the 
%%%%  journal you intend to submit to. The submission guidelines and policies 
%%%%  of the journal take precedence. A detailed User Manual is available in the 
%%%%  template package for technical guidance.
%%%%%=============================================================================%%%%

%% as per the requirement new theorem styles can be included as shown below
% \theoremstyle{thmstyleone}%
% \newtheorem{theorem}{Theorem}%  meant for continuous numbers
% %%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
% %% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
% \newtheorem{proposition}[theorem]{Proposition}% 
% %%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.

% \theoremstyle{thmstyletwo}%
% \newtheorem{example}{Example}%
% \newtheorem{remark}{Remark}%

% \theoremstyle{thmstylethree}%
% \newtheorem{definition}{Definition}%

\raggedbottom
%%\unnumbered% uncomment this for unnumbered level heads

@@

\title[MaBoSS for HPC environments]{MaBoSS for HPC environments: Implementations of the continuous time Boolean model simulator for large CPU clusters and GPU accelerators}

%%=============================================================%%
%% GivenName	-> \fnm{Joergen W.}
%% Particle	-> \spfx{van der} -> surname prefix
%% FamilyName	-> \sur{Ploeg}
%% Suffix	-> \sfx{IV}
%% \author*[1,2]{\fnm{Joergen W.} \spfx{van der} \sur{Ploeg} 
%%  \sfx{IV}}\email{iauthor@gmail.com}
%%=============================================================%%

% \author*[1,2]{\fnm{First} \sur{Author}}\email{iauthor@gmail.com}

% \author[2,3]{\fnm{Second} \sur{Author}}\email{iiauthor@gmail.com}
% \equalcont{These authors contributed equally to this work.}

% \author[1,2]{\fnm{Third} \sur{Author}}\email{iiiauthor@gmail.com}
% \equalcont{These authors contributed equally to this work.}

% \affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{100190}, \state{State}, \country{Country}}}

% \affil[2]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{10587}, \state{State}, \country{Country}}}

% \affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}

\author[1]{\fnm{Adam} \sur{Šmelko}}\email{smelko@d3s.mff.cuni.cz}

\author[2]{\fnm{Miroslav} \sur{Kratochvíl}}

\author[3,4,5]{\fnm{Emmanuel} \sur{Barillot}}
\author[3,4,5]{\fnm{Vincent} \sur{Noël}}\email{vincent.noel@curie.fr}

\affil*[1]{\orgdiv{Department of Distributed and Dependable Systems}, \orgname{Charles University}, \orgaddress{\city{Prague},  \country{Czech Republic}}}
\affil[2]{\orgdiv{Luxembourg Centre for Systems Biomedicine}, \orgname{University of Luxembourg}, \orgaddress{\city{Esch-sur-Alzette},  \country{Luxembourg}}}
\affil[3]{\orgdiv{Institut Curie}, \orgname{Université PSL, F-75005}, \orgaddress{\city{Paris},  \country{France}}}
\affil[4]{\orgdiv{INSERM}, \orgname{U900, F-75005}, \orgaddress{\city{Paris},  \country{France}}}
\affil[5]{\orgdiv{Mines ParisTech}, \orgname{Université PSL, F-75005}, \orgaddress{\city{Paris},  \country{France}}}

%%==================================%%
%% Sample for unstructured abstract %%
%%==================================%%

\abstract{Computational models in systems biology are becoming more important with the advancement of experimental techniques to query the mechanistic details responsible for leading to phenotypes of interest. In particular, Boolean models are well fit to describe the complexity of signaling networks while being simple enough to scale to a very large number of components. With the advance of Boolean model inference techniques, the field is transforming from an artisanal way of building models of moderate size to a more automatized one, leading to very large models. In this context, adapting the simulation software for such increases in complexity is crucial. 
We present two new developments in the continuous time Boolean simulators: MaBoSS.MPI, a parallel implementation of MaBoSS which can exploit the computational power of very large CPU clusters, and MaBoSS.GPU, which can use GPU accelerators to perform these simulations. }

%%================================%%
%% Sample for structured abstract %%
%%================================%%

% \abstract{\textbf{Purpose:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Methods:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Results:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.
% 
% \textbf{Conclusion:} The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. The abstract must not include subheadings (unless expressly permitted in the journal's Instructions to Authors), equations or citations. As a guide the abstract should not exceed 200 words. Most journals do not set a hard limit however authors are advised to check the author instructions for the journal they are submitting to.}

\keywords{Computational Biology, High Performance Computing, Boolean models}

%%\pacs[JEL Classification]{D8, H51}

%%\pacs[MSC Classification]{35A01, 65L10, 65L12, 65L20, 65L70}

\maketitle

@@

The source code of the proposed implementations is publicly available at their respective GitHub repositories\footnote{\url{https://github.com/sysbio-curie/MaBoSS.GPU}, \url{https://github.com/sysbio-curie/MaBoSS-env-2.0}}. We also provide the scripts, presented plots, data and instructions to reproduce the benchmarks in the replication package\footnote{\url{https://github.com/sysbio-curie/gigascience24-artifact}}.

@@

\subsubsection*{Simulation}

@@

\subsubsection*{Statistics aggregation}

@@

The MPI implementation benchmarks were run on the MareNostrum 4 supercomputer\footnote{\url{https://www.bsc.es/marenostrum/marenostrum}}.

@@

\section*{Availability and requirements}
% flex and bison are linux only :/
\begin{itemize}
	\item Project name: MaBoSS.GPU
	\item Project home page: \url{https://github.com/sysbio-curie/MaBoSS.GPU}
	\item Operating system(s): Platform independent
	\item Programming language: C++, CUDA
	\item Other requirements: Flex, Bison, CMake $>=$ 3.18, Cuda toolkit $>=$ 12.0
	\item License: MIT
\end{itemize}

@@

\backmatter

\section*{Declarations}

% Some journals require declarations to be submitted in a standardised format. Please check the Instructions for Authors of the journal to which you are submitting to see if you need to complete this section. If yes, your manuscript must contain the following sections under the heading `Declarations':

\subsection*{Availability of data and materials}

The scripts, presented plots, data and instructions to reproduce the benchmarks are available on GitHub~\cite{adam_smelko_2024_10841007}.

\subsection*{Competing Interests}
The authors declare that they have no competing interests.

\subsection*{Funding}
The research leading to these results has received funding from the European Union's Horizon 2020 Programme under the PerMedCoE Project (\url{http://www.permedcoe.eu}), grant agreement n\textsuperscript{o}~951773.
% This work was supported by the European Commission under the PerMedCoE project [H2020-ICT-951773]
The project was partially supported by Charles University, SVV project number~260698.

\subsection*{Acknowledgements}
We thank Laurence Calzone and Gautier Stoll for their guidance and fruitful discussions.

%%===================================================%%
%% For presentation purpose, we have included        %%
%% \bigskip command. Please ignore this.             %%
%%===================================================%%
% \bigskip
% \begin{flushleft}%
% Editorial Policies for:

% \bigskip\noindent
% Springer journals and proceedings: \url{https://www.springer.com/gp/editorial-policies}

% \bigskip\noindent
% Nature Portfolio journals: \url{https://www.nature.com/nature-research/editorial-policies}

% \bigskip\noindent
% \textit{Scientific Reports}: \url{https://www.nature.com/srep/journal-policies/editorial-policies}

% \bigskip\noindent
% BMC journals: \url{https://www.biomedcentral.com/getpublished/editorial-policies}
% \end{flushleft}

% \begin{appendices}

% \section{Section title of first appendix}\label{secA1}

% An appendix contains supplementary information that is not an essential part of the text itself but which may be helpful in providing a more comprehensive understanding of the research problem or it is information that is too cumbersome to be included in the body of the paper.

%%=============================================%%
%% For submissions to Nature Portfolio Journals %%
%% please use the heading ``Extended Data''.   %%
%%=============================================%%

%%=============================================================%%
%% Sample for another appendix section			       %%
%%=============================================================%%

%% \section{Example of another appendix section}\label{secA2}%
%% Appendices may be used for helpful, supporting or essential material that would otherwise 
%% clutter, break up or be distracting to the text. Appendices can consist of sections, figures, 
%% tables and equations etc.

% \end{appendices}

%%===========================================================================================%%
%% If you are submitting to one of the Nature Portfolio journals, using the eJP submission   %%
%% system, please include the references within the manuscript file itself. You may do this  %%
%% by copying the reference list from your .bbl file, paste it into the main manuscript .tex %%
%% file, and delete the associated \verb+\bibliography+ commands.                            %%
%%===========================================================================================%%

\bibliography{bibliography}% common bib file
%% if required, the content of .bbl file can be included here once bbl is generated
%%\input sn-article.bbl
